### 2020-01-02 manuscript/manuscript.rmd
add recoding and update manuscript

---
bibliography: manuscript.bib
csl: plos.csl
output:
  html_document: default
  word_document: default
  pdf_document: default
  github_document: default
---

@@

## Abstract
Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages. However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life. Here, we systematically explore the mechanisms (evolutionary models, gene sampling and outgroup choice) that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies. We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. Moreover, We find Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of the most closely related outgroups, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between site-homogenious and site-heterogeneous models. Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of current understanding of the anima-root position and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Table 1.
```{r overview data matrices}
	kable(datasets)

@@

clade_colors = viridis( length(clades) )
names( clade_colors ) = clades

ggplot(data=taxon_rectangles) +
	scale_x_continuous(name="x") +
	scale_y_continuous(name="y") +
	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
	# scale_fill_manual(guide = guide_legend(reverse=TRUE))
	# scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))
	scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
	facet_wrap( ~ matrix )

@@

Previous reanalyses have mainly focused on ourgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated. Our initial intent of comparing matrix overlap across matrices was to construct a new matrice with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap). These results indicated that different studies essentially sampled different regions of the genomes with little overlap.

@@

Here we reanalyzed `r nrow(analyses_datasets)` from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets). We also progressively trimmed each data matrix with different level of outgroup choice if applicable. We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except for large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices). We found in all cases, C60+ WAG/LG models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister in every models, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.

@@

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix. Consistent with previous study, we recovered the strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Whelan2015_D20. In contrast, we found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017. Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice will affect phylogenetic inference under CAT model. We found strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups. Thus, our results show that both model choice and outgroup sampling play a major role in incongruent results of the root position of animal phylogeny in several key data matrices.

@@

To further explore the effect of model and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices that have incongruent results with two outgroup choice, including Philippe2009, Ryan2013_est and Whelan2017_full (Fig XXPhylogenetic signal) and four different models in both ML and BI framework. By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS)in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes or sites in every analysis when using LG+GAMMA and C models and the results are consistent with best-likelihood tree always favors Ctenophora-sister over Porifera-sister in all IQtree analysis. However, we found significant differences of phylogenetic distribution between LG and C models. We found that Ctenophora-sister had the higher proportions of supporting genes when site-homogeneous models were used (52.34% to 66.51%), and the outgroup choice has little impact on the distribution of the support phylogenetic signal. This finding is consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models [@shen2017contentious].

In contrast to site-homogenious models, the distribution of phylogenetic signal over the animal-root position has never been quantified with site-heterogeneous models. Surprisingly, except for Whelan2017_Holozoa matrix, we found that the phylogenetic signal decreases in many genes using C models compared to WAG models, especially in two matrices from Ryan_2013 that over 30% of genes changed from strong to week signal (\Delta$lnL<2). It is not clear the reason why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing (control test will be conducted here, by changing monophyletic of bilaterian vs. non-monophyletic of bilaterians...).

### Bias of data-recoding

```{r phylogenetic_signal}
au_tests %>%
  unite(Matrix, c(manuscript, matrix)) %>%
  group_by(Matrix, result) %>%
  count() %>%
  group_by(Matrix) %>%
  mutate(percent = prop.table(n)) %>%

ggplot(aes(x = Matrix, y = percent, fill=result)) +
  geom_bar(stat = "identity") +
  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))

```
**Fig XXPhylogenetic signal.** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage. Proportion of genes supporting each of two alternative hypotheses for each of

@@

### Interpretting variation support

External criteria, eg posterior predictive scores, model fit etc

@@

#### phylogenetic distribution of support
To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from 3 studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices. We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta). For ML analysis in each dataset, site-wise likelihood scores ere inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG models. The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans. The constraint trees were conducted in a customized R script. The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al. 2017 [@shen2017contentious]. For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets.

#### Performance of data recoding

@@

- Six-state amino acid recoding. Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation. Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.

@@

## Model matrix comparison

WAG and LG are both fixed exchange matrices. Their differences are largely limitted to a few amino acid changes.

```{r wagvlg}

# Comparison of LG and WAG matrices

library( tidyverse )
library( magrittr )

# Order in iqtree matrix
aa = c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")

# Functional groups adjacent, from https://en.wikipedia.org/wiki/Amino_acid
aa_functional = c( "R", "H", "K", "D", "E", "S", "T", "N", "Q", "C", "G", "P", "A", "V", "I", "L", "M", "F", "Y", "W" )

triange_to_matrix = function( triangle ){
	R_matrix = rep(NA, 400)
	dim( R_matrix ) = c(20,20)

	rownames( R_matrix ) = aa
	colnames( R_matrix ) = aa

	# Though iqtree specifies lower triangle, actually in the correct order for upper triangle as this R command reads it
	R_matrix [ upper.tri( R_matrix ) ] = triangle

	# Now fill in the lower half of the matrix
	R_matrix[ is.na(R_matrix) ] = t( R_matrix )[ is.na(R_matrix) ]

	# Scale
	R_matrix = R_matrix / mean( R_matrix, na.rm=TRUE )

	# Reorder by functional groupings
	R_matrix = R_matrix[, match(aa_functional, colnames(R_matrix) )]
	R_matrix = R_matrix[match(aa_functional, rownames(R_matrix)), ]

	R_matrix
}

plot_matrix = function( R_matrix, title=NULL ){
	D = data.frame( R_matrix )
	D$source = row.names(D)
	Dg = gather( D, 1:20, key=dest, value=exch, -source )
	Dg %<>%
		mutate( source=factor(source, levels=aa_functional)  ) %>%
		mutate( dest=factor(dest, levels=aa_functional)  )

	Dg %>%
		ggplot(aes(x=dest, y=source)) +
		geom_tile(aes(fill=exch)) +
		scale_fill_gradient2() +
		ggtitle(title)
}

# Triangle matrices from iqtree source code at https://github.com/Cibiv/IQ-TREE/blob/master/model/modelprotein.cpp

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
output:
  html_document: default
  word_document: default
  pdf_document: default
  github_document: default
---

@@

## Abstract
Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages. However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life. Here, we systematically explore the mechanisms (evolutionary models, gene sampling and outgroup choice) that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies. We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. Moreover, We find Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of the most closely related outgroups, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (in site-heterogeneous CAT models) . Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of current understanding of the first diverging animal branch and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

```{r overview data matrices}
table1 = select(datasets, -c(paths))
	kable(table1, caption= "Table 1. Overview of data matrices used in this study")

@@

taxon_rectangles$matrix[taxon_rectangles$matrix == "Simion2017_subsampled"] <- "Simion2017"
taxon_rectangles$matrix[taxon_rectangles$matrix == "Hejnol2009_subsampled"] <- "Hejnol2009"

clade_colors = c("#e31a1c","#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#a6cee3","#fdbf6f","#cab2d6", "#ff7f00")
names( clade_colors ) = clades

ggplot(data=taxon_rectangles) +
	scale_x_continuous(name="Gene sampling") +
	scale_y_continuous(name="Taxa sampling") +
	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
	#scale_fill_manual(guide = guide_legend(reverse=TRUE))+
 scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))+
	#scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
	facet_wrap( ~ matrix )

@@

Previous reanalyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated. Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap). These results indicated that different studies essentially sampled different regions of the genomes with little overlap.

@@

Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets). We also progressively trimmed each data matrix with different level of outgroup choice if applicable. We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except for large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices). We found in all cases, C60+ WAG/LG models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister in every models, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.

@@

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix. Consistent with previous study, we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Nosenko2013 and Whelan2015. In contrast, we found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017. Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice will affect phylogenetic inference under CAT model. We found strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups. Thus, our results show that both model choice and outgroup sampling play a major role in incongruent results of the root position of animal phylogeny in several key data matrices.

@@

The distribution of phylogenetic signal over the root position of animal phylogeny has never been quantified with site-heterogeneous models. To further explore the effect of model selection and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with two outgroup choice and four models in both ML and BI framework (Fig XXPhylogenetic signal). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models. Moreover, the outgroup choice has little impact in general on the distribution of the support phylogenetic signal in site-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

In contrast, the distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in two matrices from Ryan_2013 that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. phylogenetic_signal B). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of root position of animal phylogeny. In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT model and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal. We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled genes) when exclusive of more distantly related outgroups with the Poisson-CAT models. The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of genes that are supported in each hypothesis across different analyses in Whelan2017 matrices. Consistent with the summary of \Delta$GLS analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of the root of animal phylogeny is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in these data matrices. Thus, it is clear that recovering Porifera-sister requires more constrains including model, outgroup choice, and even with gene sampling.

```{r phylogenetic_signal}
#au_tests %>%
#  unite(Matrix, c(manuscript, matrix, model)) %>%
#  group_by(Matrix, result) %>%
#  count() %>%
#  group_by(Matrix) %>%
#  mutate(percent = prop.table(n)) %>%

#ggplot(aes(x = Matrix, y = percent, fill=result)) +
#  geom_bar(stat = "identity") +
#  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
#  theme(axis.text.x = element_text(angle=45, hjust=1)) +
#  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))

library(ggpubr)
theme_set(theme_pubr())
au_tests = read_tsv("../trees_new/AU_test/summary.tsv")

 Philippe2009 = filter(au_tests,
                   grepl('Philippe', Matrix))

x1  = factor(Philippe2009$Matrix, levels=c("Philippe2009_WAG", "Philippe2009_Choanozoa_WAG", "Philippe2009_LG", "Philippe2009_Choanozoa_LG", "Philippe2009_LG(phylobayes)", "Philippe2009_Choanozoa_LG(phylobayes)", "Philippe2009_C60", "Philippe2009_Choanozoa_C60", "Philippe2009_F81", "Philippe2009_Choanozoa_F81"))
y1=factor(Philippe2009$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))

 Ryan2013_est = filter(au_tests,
                   grepl('Ryan', Matrix))

y2=factor(Ryan2013_est$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))

x2  = factor(Ryan2013_est$Matrix, levels=c("Ryan2013_est_WAG", "Ryan2013_Choanozoa_WAG", "Ryan2013_est_LG", "Ryan2013_Choanozoa_LG",  "Ryan2013_est_C60", "Ryan2013_Choanozoa_C60"))

Whelan2017 = filter(au_tests,
                   grepl('Whelan', Matrix))
y3=factor(Whelan2017$result, levels=c("Ctenophore-sister-strong-signal", "Ctenophore-sister-weak-signal", "Porifera-sister-weak-signal", "Porifera-sister-strong-signal"))

x3  = factor(Whelan2017$Matrix, levels=c("Whelan2017_full_WAG", "Whelan2017_Choanozoa_WAG", "Whelan2017_full_LG", "Whelan2017_Choanozoa_LG", "Whelan2017_full_LG(phylobayes)", "Whelan2017_Choanozoa_LG(phylobayes)", "Whelan2017_full_C60", "Whelan2017_Choanozoa_C60", "Whelan2017_full_F81", "Whelan2017_Choanozoa_F81"))

p1 <- ggplot(Philippe2009, aes(x = x1, y = percent, fill=y1)) +
  geom_bar(stat = "identity") +
  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))

p2 <- ggplot(Ryan2013_est, aes(x = x2, y = percent, fill=y2)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))

p3 <- ggplot(Whelan2017, aes(x = x3, y = percent, fill=y3)) +
  geom_bar(stat = "identity") +
  #scale_y_continuous(limits = c(0,1), expand = c(0, 0))+
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  scale_fill_manual(values=c("#E69F00", "#ebd196","#a2e8d5","#009E73"))

figure <- ggarrange(p1, p2, p3,
                    labels = c("A", "B", "C"),
                    ncol = 1, nrow = 3, widths=c(2,1,2), heights=c(1,1,1), common.legend = TRUE, legend = "right")
```

### Bias of data-recoding
There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies. They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices. Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Figure ). Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Figure 1B).Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).

**Fig XXPhylogenetic signal.** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage. Proportion of genes supporting each of two alternative hypotheses for each of

@@

@@

#### phylogenetic distribution of support
To investigate the distribution of phylogenetic signal in data matrices, we considered three major data matrices from 3 studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013, and Whelan_2017 data matrices. We examined two hypotheses: Ctenophora-sister and Porifera-sister to rest of metazoans, under both ML and BI frameworks with different outgroup schemes (Choanomalia and Opisthokonta). For ML analysis in each dataset, site-wise likelihood scores were inferred for both hypotheses using IQtree (option -g) with the same best-fit model identified above, and site-homogeneous WAG models. The two different phylogenetic hypotheses passed to IQtree (via -z) were the corresponding tree that the ctenophore as the sister lineage tree and the corresponding tree that was modified to have sponges as the sister to all other metazoans. The constraint trees were conducted in a customized R script. The numbers of genes and sites supporting each hypothesis were calculated with IQtree output and Perl scripts from Shen et al. 2017 [@shen2017contentious]. For BI analysis, we only considered the Philippe_2009 and Whelan_2017 datasets.

#### Performance of data-recoding
All code used for the analyses presented here is available in a git repository at https://github.com/caseywdunn/feuda_2017_response. The randomized recoding analyses are in the `../trees_new/recoding/feuda_2017_response/alternative` subdirectory of the git repository.

The original SR-6 recoding scheme is "`APST CW DEGN FHY ILMV KQR`" [@Susko:2007ds], where spaces separate amino acids that are placed into the same group. This recoding is one member of a family of recodings, each with a different number of groups, based on clustering of the JTT matrix. The other recoding used by Feuda *et al.*, KGB-6 and D-6, are based on different matrices and methods [@Feuda:2017ew].

The `alt_recode.py` script was used to generate the randomized recoding schemes and apply the recodings to the data. To create the randomized recoding schemes, the amino acids in the SR-6 recoding were randomly reshuffled. This creates new recodings that have the same sized groups as SR-6. The new recodings were, from `random-00` to `random-03`:

    MPKE AY IDGQ TRC WNLF SVH
    EIFT WL QVPG NKM SCHA RYD
    LCQS GK WPTI VRD YEFN MAH
    IWQV TY KDLM ESH APCF GRN

To apply these to the data, each amino acid was replaced with the first amino acid in the group. When applying `random-00`, for example, each instance of `R` and `C` would be replaced by a `T`.

The 20 state matrices are the same across all analyses since they are not recoded. Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Figure 1B) give insight into the technical variance of the inference process.

Each new matrix was analyzed with PhyloBayes-mpi. Analyses were run for 1000 generations, and a 200 generation burnin applied. The resulting tree files and posterior predictive scores were parsed for display with the code in `manuscript.rmd`.

The statistics presented in Supplemental Figures 1 and 2 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.

@@

- Recoding method:  Amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation. Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.

@@

## Data-recoding methods

Feuda *et al.* [@Feuda:2017ew] tackled a difficult phylogenetic problem -- placing the root of the animal phylogeny [@King:2017ie]. In the past decade, some analyses have placed Ctenophora (comb jellies) and others Porifera (sponges) as the sister group to all other animals (Fig ##Random_recodingA). Feuda *et al.* [@Feuda:2017ew] present new phylogenetic analyses that they claim provide strong support for Porifera-sister. Their analyses consider datasets from studies by Chang *et al.* [@Chang:2015hl] and Whelan *et al.* [@Whelan:2015jj], both of which found support for Ctenophora-sister. Feuda *et al.* were concerned that these Ctenophora-sister results were artefacts of lineage-specific differences in amino acid frequencies. In an attempt to reduce these differences, they recoded the full set of twenty amino acids into six groups of amino acids. These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [@Susko:2007ds]. The intent is to discard many lineage-specific changes, which are expected to fall within these groups. Rather than model compositional heterogeneity, as their title suggests, this approach discards heterogeneous information so that much simpler models with fewer states can be applied. They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies". Here we focus on two aspects of Feuda *et al.* First, we point out that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister. Second, we present new analyses that show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister. They also show that recoding can be a problematic method for addressing compositional variation.

Feuda *et al.* examine support for Ctenophora-sister and Porifera-sister under all combinations of two models of molecular evolution, four datasets, and four coding schemes. This provides 32 analyses that they report in their Table 3 and that we present graphically here as Supplemental Figure ##recoding_summary. There is striking variation in support for Ctenophora-sister and Porifera-sister across these analyses (Figure ##recoding_summary). Feuda *et al.* accept the results of some analyses and reject others based on posterior predictive (PP) analyses of model adequacy, which assess how well a model explains variation in the data [@Bollback:2002to]. They considered five different posterior predictive statistics that capture different types of variation in the data. From this they conclude that their "results strongly support sponges as the sister group of all other animals".

This conclusion does not follow from their own presented results. Only a single analysis with posterior predictive scores provides what could be considered strong support > 95 posterior probability) for Porifera-sister. Of the 32 analyses, posterior predictive scores were calculated for 16 (those for the full Whelan and Chang matrices). Based on posterior predictive scores, Feuda *et al.* reject eight of these that were conducted under the GTR+G model (which all have strong support for Ctenophora-sister). This leaves eight CAT-GTR+G analyses (Figure ##recoding_summary). Two of these eight are analyses of the original 20-state amino acid data, both of which provide strong support for Ctenophora-sister. Of the six recoded analyses, five are unresolved. Only a single analysis for which posterior predictive scores are available provides strong support for Porifera-sister, the CAT-GTR+G analysis of the SR-6 [@Susko:2007ds] recoded Whelan [@Whelan:2015jj] matrix. Furthermore, this analysis does not have the best score according to any of the five posterior predictive statistics they considered (Figure ##recoding_summary). The only statistic that stands out for this one analysis is that it has the highest maxdiff (Figure ##recoding_summary), indicating that it did not converge as well as other analyses.

Though their study does not provide strong support for Porifera-sister, the sensitivity of their results to recoding provides an opportunity to better understand and evaluate the impact of recoding more generally. This is important given the growing interest in recoding [@Feuda:2017ew]. Feuda *et al.* hoped recoding would reduce potential artefacts due to differences across species in amino acid frequencies. They interpreted the fact that their analyses are sensitive to recoding as evidence that such an artefact exists and that they successfully addressed it by recoding. An alternative hypothesis is that recoding impacts phylogenetic analyses because it discards so much information. These two hypotheses can be tested by applying new recoding schemes that also reduce twenty states down to six, but are based on random grouping rather than empirical frequencies of amino acid exchange. Empirical and random recodings both discard the same amount of information, but only empirical recoding reduce the impact of amino-acid frequency as intended. Different results between empirical and random recoding would be consistent with the hypothesis that the empirical approach works as intended to accommodate compositional heterogeneity. Similar results would suggest that the impact of recoding is due to discarding information. Here we focus on their single analysis with a posterior predictive score that supports Porifera-sister, the CAT-GTR+G analysis of the SR-6 recoded Whelan data. we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see Supplemental Methods and analysis code at https://github.com/caseywdunn/feuda_2017). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi [@Lartillot:2013fg], we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Fig ##random_recodingB).

These analyses suggest that the major impact of recoding on phylogenetic analyses is data reduction, not accommodation of compositional heterogeneity across species. This indicates that recoding may not be an effective tool for accommodating among-species differences in amino acid frequencies. Compositional heterogeneity would be better addressed with models of molecular evolution that explicitly describe such differences [@Blanquart:2008gl, @Foster:2004tw], if progress can be made on the considerable computational challenges of such complex models.

![](figures/Figure_overview.png)
**Fig ##Random_recoding.**  A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3. Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved). Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
